---
title: "Chess Learning — Weight Trajectories & Learning Analysis"
subtitle: "Stage 03 Visualization"
author: "Posit AI, under guidance from Dan Killian"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    fig-height: 4
    fig-length: 6
execute:
  warning: false
  message: false
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F, fig.height=3, fig.width=5}
library(here)
library(knitr)
library(DBI)
library(RSQLite)
library(flextable)
source("../../../prep (May 2025).r")
theme_set(base_quarto)
```


```{r setup}
# Path to database — works when rendered from scripts/r/ or run interactively
DB_PATH <- if (file.exists("../../data/chess_learning.db")) {
  "../../data/chess_learning.db"
} else {
  "data/chess_learning.db"
}

# Feature groupings — 17 features across 4 groups
MATERIAL_FEATURES    <- c("material_queen", "material_rook", "material_bishop",
                           "material_knight", "material_pawn")
PAWN_FEATURES        <- c("passed_pawn", "doubled_pawn", "isolated_pawn", "pawn_advancement")
POSITIONAL_FEATURES  <- c("king_safety", "center_control", "rook_open_file",
                           "connected_rooks", "rook_seventh")
DEVELOPMENT_FEATURES <- c("bishop_pair", "piece_development", "mobility")

ALL_GROUPS <- c("Material", "Pawn structure", "Positional", "Coordination & development")

FEATURE_ORDER <- c(MATERIAL_FEATURES, PAWN_FEATURES, POSITIONAL_FEATURES, DEVELOPMENT_FEATURES)

FEATURE_GROUP <- tibble(
  feature_name = FEATURE_ORDER,
  group = c(
    rep("Material",                   length(MATERIAL_FEATURES)),
    rep("Pawn structure",             length(PAWN_FEATURES)),
    rep("Positional",                 length(POSITIONAL_FEATURES)),
    rep("Coordination & development", length(DEVELOPMENT_FEATURES))
  )
)

# Canonical piece values (for reference lines in material section)
CANONICAL <- tibble(
  feature_name  = MATERIAL_FEATURES,
  canonical_val = c(9, 5, 3, 3, 1)
)

# Outcome → numeric from white's perspective
outcome_numeric <- function(outcome) {
  case_when(
    outcome == "1-0"     ~  1,
    outcome == "0-1"     ~ -1,
    outcome == "1/2-1/2" ~  0,
    .default = NA_real_
  )
}

# Rolling correlation helper
roll_cor <- function(x, y, window = 100) {
  n <- length(x)
  map_dbl(seq_len(n), function(i) {
    if (i < window) return(NA_real_)
    cor(x[(i - window + 1):i], y[(i - window + 1):i], use = "complete.obs")
  })
}

# Connect and load tables
con    <- dbConnect(SQLite(), DB_PATH)
runs   <- dbReadTable(con, "runs")
games  <- dbReadTable(con, "games")

# Deduplicate weights: keep the last-written row per (run_id, game_number, feature_name)
weights <- dbGetQuery(con,
  "SELECT run_id, game_number, feature_name, weight_value
   FROM weights
   GROUP BY run_id, game_number, feature_name
   HAVING rowid = MAX(rowid)"
)

has_feature_stats <- dbExistsTable(con, "feature_stats")
has_game_records  <- dbExistsTable(con, "game_records")
has_weight_deltas <- dbExistsTable(con, "weight_deltas")

feature_stats <- if (has_feature_stats) dbReadTable(con, "feature_stats") else NULL
game_records  <- if (has_game_records)  dbReadTable(con, "game_records")  else NULL
weight_deltas <- if (has_weight_deltas) dbReadTable(con, "weight_deltas") else NULL
dbDisconnect(con)

run_ids <- sort(unique(games$run_id))
cat("Runs in database:", paste(run_ids, collapse = ", "), "\n")
games |> count(run_id) |> print()
```

## Weight Trajectories

How each feature weight evolves over the course of training.
Features are grouped by type. Weights start near zero and should drift
toward values that reflect the true importance of each chess concept.

```{r weight-trajectories}
weights_tidy <- weights |>
  left_join(FEATURE_GROUP, by = "feature_name") |>
  mutate(
    feature_name = factor(feature_name, levels = FEATURE_ORDER),
    group        = factor(group, levels = ALL_GROUPS)
  )

plot_weight_group <- function(grp) {
  weights_tidy |>
    filter(group == grp) |>
    ggplot(aes(x = game_number, y = weight_value, colour = feature_name)) +
    geom_line(linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    facet_wrap(~ run_id, labeller = label_both, scales = "free_x") +
    labs(
      title  = paste(grp, "weights over training"),
      x      = "Game number",
      y      = "Weight value",
      colour = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

plot_weight_group("Material")
plot_weight_group("Pawn structure")
plot_weight_group("Positional")
plot_weight_group("Coordination & development")
```

### Material weights vs canonical values

Canonical piece values (queen=9, rook=5, bishop=3, knight=3, pawn=1) normalised to
pawn=1. This panel shows whether the *relative* ordering of material weights is
converging toward the accepted hierarchy.

```{r material-vs-canonical}
weights |>
  filter(feature_name %in% MATERIAL_FEATURES) |>
  pivot_wider(names_from = feature_name, values_from = weight_value) |>
  filter(!is.na(material_pawn), material_pawn != 0) |>
  mutate(across(all_of(MATERIAL_FEATURES), ~ . / abs(material_pawn))) |>
  pivot_longer(all_of(MATERIAL_FEATURES),
               names_to = "feature_name", values_to = "normalised_weight") |>
  left_join(CANONICAL, by = "feature_name") |>
  ggplot(aes(x = game_number, y = normalised_weight, colour = feature_name)) +
  geom_line(linewidth = 0.8) +
  geom_hline(
    data = CANONICAL,
    aes(yintercept = canonical_val, colour = feature_name),
    linetype = "dotted", linewidth = 0.6
  ) +
  facet_wrap(~ run_id, labeller = label_both, scales = "free_x") +
  labs(
    title  = "Material weights normalised to pawn (dotted = canonical value)",
    x      = "Game number",
    y      = "Weight relative to |pawn weight|",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Game Statistics

How the character of games changes as the engine learns.

```{r game-stats}
games_clean <- games |>
  mutate(outcome_label = case_when(
    outcome == "1-0"     ~ "White wins",
    outcome == "0-1"     ~ "Black wins",
    outcome == "1/2-1/2" ~ "Draw",
    .default = "Other"
  ))

WIN <- 50

games_rolling <- games_clean |>
  arrange(run_id, game_number) |>
  group_by(run_id) |>
  mutate(
    white_win  = (outcome == "1-0"),
    draw       = (outcome == "1/2-1/2"),
    roll_white = slider::slide_dbl(white_win,   mean, .before = WIN - 1, .complete = FALSE),
    roll_draw  = slider::slide_dbl(draw,        mean, .before = WIN - 1, .complete = FALSE),
    roll_len   = slider::slide_dbl(n_halfmoves, mean, .before = WIN - 1, .complete = FALSE)
  ) |>
  ungroup()
```

```{r outcome-rates, fig.height=4}
games_rolling |>
  select(run_id, game_number, roll_white, roll_draw) |>
  pivot_longer(c(roll_white, roll_draw),
               names_to = "series", values_to = "rate") |>
  mutate(series = if_else(series == "roll_white", "White win rate", "Draw rate")) |>
  ggplot(aes(x = game_number, y = rate, colour = series)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ run_id, labeller = label_both, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title  = paste0("Rolling outcome rates (", WIN, "-game window)"),
    x      = "Game number",
    y      = "Rate",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r game-length, fig.height=4}
games_rolling |>
  ggplot(aes(x = game_number, y = roll_len)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~ run_id, labeller = label_both, scales = "free_x") +
  labs(
    title = paste0("Rolling average game length (", WIN, "-game window)"),
    x     = "Game number",
    y     = "Half-moves per game"
  ) +
  theme_minimal(base_size = 12)
```

```{r termination-type, fig.height=4}
games_clean |>
  count(run_id, terminated_by) |>
  group_by(run_id) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = terminated_by, y = pct, fill = terminated_by)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)), vjust = -0.3, size = 3.5) +
  facet_wrap(~ run_id, labeller = label_both) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    title = "Game termination type by run",
    x     = NULL,
    y     = "Share of games"
  ) +
  theme_minimal(base_size = 12)
```

## Most Impactful Games

Which individual games drove the largest weight change for a given feature?
Because temporal difference learning updates weights after every game, some games — depending on their
outcome, length, and the positions reached — produce outsized shifts. Identifying these
games lets you examine the actual play that taught the engine a concept.

Set `FOCUS_FEATURE` and `TOP_N` to explore any feature.

```{r impactful-games-setup}
FOCUS_FEATURE <- "material_queen"
TOP_N         <- 10
```

```{r impactful-games-table}
if (!has_game_records || !has_weight_deltas) {
  cat(
    "game_records and weight_deltas tables not yet populated.\n",
    "Run selfplay.ipynb to generate this data.\n"
  )
} else {
  top_games <- weight_deltas |>
    filter(feature_name == FOCUS_FEATURE) |>
    mutate(abs_delta = abs(delta)) |>
    arrange(desc(abs_delta)) |>
    slice_head(n = TOP_N) |>
    left_join(
      game_records |> select(run_id, game_number, pgn, td_error),
      by = c("run_id", "game_number")
    ) |>
    left_join(
      games |> select(run_id, game_number, outcome, n_halfmoves),
      by = c("run_id", "game_number")
    ) |>
    mutate(
      moves = map_chr(pgn, \(p) {
        parts <- str_split(p, "\n\n")[[1]]
        tail(parts, 1) |> str_squish() |> str_trunc(80)
      })
    ) |>
    select(run_id, game_number, outcome, n_halfmoves, delta, td_error, moves)

  top_games |>
    mutate(across(where(is.double), ~ round(., 4))) |>
    flextable() |>
    set_header_labels(
      run_id      = "Run",
      game_number = "Game",
      outcome     = "Outcome",
      n_halfmoves = "Half-moves",
      delta       = paste0("\u0394 ", FOCUS_FEATURE),
      td_error    = "TD error",
      moves       = "Moves (truncated)"
    ) |>
    flextable::set_caption(paste0("Top ", TOP_N, " games by |\u0394 ", FOCUS_FEATURE, "|")) |>
    autofit()
}
```

```{r impactful-games-plot, fig.height=4}
if (has_game_records && has_weight_deltas) {
  delta_series <- weight_deltas |>
    filter(feature_name == FOCUS_FEATURE) |>
    left_join(
      games |> select(run_id, game_number, outcome),
      by = c("run_id", "game_number")
    ) |>
    mutate(is_top = game_number %in% top_games$game_number)

  ggplot(delta_series, aes(x = game_number, y = delta)) +
    geom_line(colour = "grey75", linewidth = 0.4) +
    geom_point(
      data = filter(delta_series, is_top),
      aes(colour = outcome),
      size = 2.5
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    facet_wrap(~ run_id, labeller = label_both, scales = "free_x") +
    labs(
      title    = paste0("Per-game weight delta: ", FOCUS_FEATURE),
      subtitle = paste0("Top ", TOP_N, " most impactful games highlighted"),
      x        = "Game number",
      y        = "\u0394 weight",
      colour   = "Outcome"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

### Per-game deltas — all features

Weight change each game across all features simultaneously. Each bar is one game;
colour shows the outcome. Games with near-zero deltas (mostly draws at early stages)
will appear flat — making decisive games visually obvious.

```{r delta-all-features, fig.height=10}
if (has_weight_deltas) {
  weight_deltas |>
    filter(run_id %in% run_ids) |>
    left_join(
      games |> select(run_id, game_number, outcome),
      by = c("run_id", "game_number")
    ) |>
    mutate(feature_name = factor(feature_name, levels = FEATURE_ORDER)) |>
    ggplot(aes(x = game_number, y = delta, fill = outcome)) +
    geom_col(width = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    facet_wrap(~ feature_name, scales = "free_y", ncol = 4) +
    labs(
      title    = "Per-game weight delta — all features",
      subtitle = "Each bar = one game; colour = outcome",
      x        = "Game number",
      y        = "\u0394 weight",
      fill     = "Outcome"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("weight_deltas table not yet populated.\n")
}
```

## Feature-Outcome Correlation

For each feature, does its average value during a game predict whether white wins?
A positive correlation means: higher average feature value (from white's perspective)
is associated with white winning.

This is the primary tool for testing the **pawn U-curve hypothesis**: if the pawn
feature-outcome correlation dips negative during the phase when queen weights are
rising, and then recovers, that is evidence of a structural learning sequence rather
than noise.

```{r feature-outcome-correlation}
if (!has_feature_stats || nrow(feature_stats) == 0) {
  cat("No feature_stats data yet. Run a training session with selfplay.ipynb",
      "to populate this table.\n")
} else {
  ROLL_WIN <- 200

  corr_data <- feature_stats |>
    left_join(FEATURE_GROUP, by = "feature_name") |>
    mutate(
      outcome_num  = outcome_numeric(outcome),
      feature_name = factor(feature_name, levels = FEATURE_ORDER),
      group        = factor(group, levels = ALL_GROUPS)
    ) |>
    arrange(run_id, feature_name, game_number) |>
    group_by(run_id, feature_name, group) |>
    mutate(roll_corr = roll_cor(mean_value, outcome_num, window = ROLL_WIN)) |>
    ungroup()

  corr_data |>
    ggplot(aes(x = game_number, y = roll_corr, colour = feature_name)) +
    geom_line(linewidth = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    facet_grid(group ~ run_id, scales = "free_x") +
    labs(
      title    = paste0("Rolling feature-outcome correlation (", ROLL_WIN, "-game window)"),
      subtitle = "Positive = higher feature value predicts white wins",
      x        = "Game number",
      y        = "Correlation with outcome",
      colour   = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## Cross-Run Comparison

Concept emergence timing across independent runs — the central research question.
A concept is considered "emerged" when its weight crosses a defined threshold for
the first time.

```{r concept-emergence}
EMERGENCE_THRESHOLD <- 0.1

emergence <- weights |>
  left_join(FEATURE_GROUP, by = "feature_name") |>
  filter(weight_value > EMERGENCE_THRESHOLD) |>
  group_by(run_id, feature_name, group) |>
  summarise(emerged_at = min(game_number), .groups = "drop") |>
  mutate(
    feature_name = factor(feature_name, levels = FEATURE_ORDER),
    group        = factor(group, levels = ALL_GROUPS)
  )

if (nrow(emergence) == 0) {
  cat("No features have yet crossed the emergence threshold (", EMERGENCE_THRESHOLD, ").\n",
      "Run more training games.\n")
} else {
  emergence |>
    ggplot(aes(x = emerged_at, y = feature_name, colour = factor(run_id))) +
    geom_point(size = 3) +
    facet_wrap(~ group, scales = "free_y", ncol = 1) +
    labs(
      title    = paste0("Concept emergence (weight > ", EMERGENCE_THRESHOLD, ")"),
      subtitle = "Each dot = first checkpoint where weight crossed threshold",
      x        = "Game number",
      y        = NULL,
      colour   = "Run"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## Final Weights Summary

Snapshot of weights at the end of each run.

```{r final-weights-table}
weights |>
  group_by(run_id) |>
  filter(game_number == max(game_number)) |>
  ungroup() |>
  left_join(FEATURE_GROUP, by = "feature_name") |>
  left_join(CANONICAL, by = "feature_name") |>
  mutate(
    feature_name = factor(feature_name, levels = FEATURE_ORDER),
    group        = factor(group, levels = ALL_GROUPS)
  ) |>
  arrange(run_id, group, feature_name) |>
  select(run_id, group, feature_name, weight_value, canonical_val) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  flextable() |>
  set_header_labels(
    run_id       = "Run",
    group        = "Group",
    feature_name = "Feature",
    weight_value = "Final weight",
    canonical_val = "Canonical"
  ) |>
  flextable::set_caption("Final weights vs canonical piece values (material features only)") |>
  autofit()
```
