---
title: "How the Chess Engine Chooses a Move"
subtitle: "Negamax search with alpha-beta pruning"
author: "Dan Killian, with Posit AI"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    fig-height: 4
    fig-width: 7
    mermaid-format: js
execute:
  warning: false
  message: false
  error: false
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(DBI)
library(RSQLite)
source("../../../prep (May 2025).R")
theme_set(base_quarto)

engine  <- reticulate::import_from_path("engine", path = "../python")
chess_m <- reticulate::import("chess")
```

## Overview

The chess engine has two jobs. The first job — covered in the TD
learning tutorial — is **evaluation**: given any board position, assign
a score representing how good it looks for the side to move.

The second job is **move selection**: given 20 or 30 legal options, pick
the best one. Evaluation alone can't do this. The move that immediately
looks best (high score after one move) might walk into a trap the
opponent springs on the very next turn. The engine needs to look
further ahead.

By the end of this tutorial you should understand:

-   Why a one-move horizon is not enough to play chess well
-   How the game tree represents all possible sequences of moves
-   How the negamax algorithm searches that tree
-   How alpha-beta pruning eliminates most of the search without
    changing the result
-   How this plays out on a real chess position

---

## Why One Move Ahead Isn't Enough

Consider the board after **1. e4 e5 2. Bc4 Nc6** — White is looking for
a move. Suppose White's evaluation function slightly favors positions
where the queen is active. Playing **3. Qh5** gives the queen a
prominent square and might score well in a one-move evaluation.

But the *reason* Qh5 is strong is not the queen's immediate position —
it's the threat on f7 that follows. If Black blunders with 3... Nf6??,
then 4. Qxf7# is checkmate. A one-move evaluation can't see that. It
takes three half-moves of lookahead (White plays, Black responds, White
delivers checkmate) to detect the threat.

The engine handles this by building a **game tree**: a branching
structure of all positions reachable over the next several moves.

---

## The Game Tree

Each node in the tree is a board position. The root is the current
position. Edges are legal moves. The tree fans out to some fixed
**depth** — in this project, depth = 3.

```{mermaid}
graph TD
    R["<b>Root</b> — White to move<br/>maximize · returns <b>5</b> · picks Move A"]

    A["<b>Move A</b> — Black to move<br/>minimize · returns <b>5</b>"]
    B["<b>Move B</b> — Black to move<br/>minimize · returns <b>4</b>"]

    A1["A → White to move<br/>maximize · returns <b>5</b>"]
    A2["A → White to move<br/>maximize · returns <b>9</b>"]
    B1["B → White to move<br/>maximize · returns <b>4</b>"]
    B2["B → White to move<br/>maximize · returns <b>7</b>"]

    LA1["3"]
    LA2["5"]
    LA3["9"]
    LA4["1"]
    LB1["2"]
    LB2["4"]
    LB3["7"]
    LB4["6"]

    R --> A & B
    A --> A1 & A2
    B --> B1 & B2
    A1 --> LA1 & LA2
    A2 --> LA3 & LA4
    B1 --> LB1 & LB2
    B2 --> LB3 & LB4
```

Leaf nodes (bottom row) are evaluated by the evaluation function — they
have no children, so there is no more lookahead. Their scores are in
White's favour if positive. Interior nodes are not evaluated directly;
instead, their score **backs up** from their children:

-   At **White's turn** (depth 3, 1), White picks the child with the
    **highest** score.
-   At **Black's turn** (depth 2), Black picks the child with the
    **lowest** score — Black is trying to minimise White's advantage.

Working from the bottom up in the diagram:

| Node | Calculation | Value |
|------|-------------|-------|
| A1   | max(3, 5)   | 5     |
| A2   | max(9, 1)   | 9     |
| A    | min(5, 9)   | 5     |
| B1   | max(2, 4)   | 4     |
| B2   | max(7, 6)   | 7     |
| B    | min(4, 7)   | 4     |
| Root | max(5, 4)   | **5** — picks Move A |

White chooses Move A because it leads to the best outcome assuming both
sides play optimally.

::: {.callout-note}
This diagram uses a branching factor of 2 for readability. A real chess
position has 20–30 legal moves per side. At depth 3, the full tree
would have up to 30³ = **27,000 leaf nodes**.
:::

---

## Negamax: Always Maximise

The minimax algorithm above has two cases: White maximises, Black
minimises. **Negamax** collapses these into a single rule: *always
maximise, but negate the child's score when returning it.*

The key insight is that what is good for the opponent is bad for you.
If a position has value +5 for the player who just moved, it has value
−5 for the player who must respond. So after each move, we flip the
sign.

```python
def negamax(board, depth, alpha, beta, weights):
    if board.is_game_over():
        return -10_000.0 if board.is_checkmate() else 0.0
    if depth == 0:
        return evaluate(board, weights)   # score for the side to move

    max_score = float("-inf")
    for move in board.legal_moves:
        board.push(move)
        score = -negamax(board, depth - 1, -beta, -alpha, weights)
        board.pop()
        if score > max_score:
            max_score = score
        if score > alpha:
            alpha = score
        if alpha >= beta:
            break
    return max_score
```

The `evaluate()` call at depth = 0 returns a score **from the
perspective of the side to move** — positive means good for whoever is
about to play. The negation on the recursive call (`-negamax(...)`)
performs the perspective flip: the score returned by the child is from
the child's side's view, so we negate it to get the parent's view.

---

## Alpha-Beta Pruning

At depth 3 with branching factor 30, the naive tree has ~27,000 leaves.
Alpha-beta pruning eliminates branches that cannot affect the final
answer.

Two parameters track the search:

-   **α (alpha)**: the best score the current player can *guarantee* so
    far (the floor).
-   **β (beta)**: the best score the *opponent* can guarantee — if our
    score exceeds this, the opponent will never let us reach this branch
    (the ceiling).

When **α ≥ β**, the branch is abandoned. There are two ways this
happens:

**Beta cutoff (inside A2 below):** White is searching A2 and finds a
leaf scoring 9. Black (the A node) already knows it can hold White to 5
via A1. Any further search in A2 can only help White — but Black would
never choose A2 anyway. Remaining A2 children are skipped.

**Alpha cutoff (B2 below):** After fully evaluating B1, Black's best
return from Move B is ≤ 4. White's α is already 5 (from Move A). White
would never choose Move B regardless of what B2 contains. B2 is skipped
entirely.

```{mermaid}
graph TD
    R["<b>Root</b> — White · α=−∞, β=+∞<br/>returns <b>5</b>, α → 5 · picks Move A"]

    A["<b>Move A</b> — Black · α=−∞, β=+∞<br/>β → 5 after A1 · returns <b>5</b>"]
    B["<b>Move B</b> — Black · α=5 inherited<br/>β → 4 after B1 · α ≥ β · pruned!"]

    A1["A1 · White<br/>returns <b>5</b>"]
    A2["A2 · White · β=5<br/>9 ≥ β · cutoff!<br/>returns ≥ 9"]
    B1["B1 · White · α=5<br/>returns <b>4</b>"]
    B2(["✗  B2 — never explored<br/>α=5 ≥ β=4"])

    LA1["3 ✓"]
    LA2["5 ✓"]
    LA3["9 ✓ → cutoff"]
    LA4(["1 ✗  skipped"])
    LB1["2 ✓"]
    LB2["4 ✓"]
    LB3(["7 ✗  skipped"])
    LB4(["6 ✗  skipped"])

    R --> A & B
    A --> A1 & A2
    B --> B1
    B -.-> B2
    A1 --> LA1 & LA2
    A2 --> LA3
    A2 -.-> LA4
    B1 --> LB1 & LB2
    B2 -.-> LB3 & LB4
```

Result: **5 of 8 leaves** were evaluated (leaves 1, 7, 6 skipped — 37%
savings). In practice, with 30 legal moves per side and depth = 3, good
move ordering reduces evaluations from ~27,000 to roughly **1,700** —
a 16× speedup — without changing the result.

::: {.callout-tip}
## Move ordering matters

Alpha-beta is most effective when strong moves are searched first,
because a high α is established early and prunes more aggressively.
The engine uses `randomize=True` in `get_best_move()` to break ties
randomly, which also prevents repetitive openings. A production engine
would use more sophisticated ordering (e.g. captures first, then checks).
:::

---

## The Code

Here are the two search functions from `engine.py`. `negamax()` does
the recursive tree search; `get_best_move()` calls it for every root
move and returns the best one.

```python
def negamax(board, depth, alpha, beta, weights):
    """
    Negamax search with alpha-beta pruning.
    Returns the score from the perspective of the side to move.
    """
    if board.is_game_over():
        return -10_000.0 if board.is_checkmate() else 0.0

    if depth == 0:
        return evaluate(board, weights)

    max_score = float("-inf")
    for move in board.legal_moves:
        board.push(move)
        score = -negamax(board, depth - 1, -beta, -alpha, weights)
        board.pop()

        if score > max_score:
            max_score = score
        if score > alpha:
            alpha = score
        if alpha >= beta:
            break   # Beta cutoff — prune remaining moves

    return max_score


def get_best_move(board, weights, depth=2, randomize=True):
    """
    Return (best_move, score) for the current side to move.
    randomize=True shuffles the move list so ties are broken randomly.
    """
    moves = list(board.legal_moves)
    if randomize:
        random.shuffle(moves)

    best_move  = None
    best_score = float("-inf")
    alpha      = float("-inf")
    beta       = float("inf")

    for move in moves:
        board.push(move)
        score = -negamax(board, depth - 1, -beta, -alpha, weights)
        board.pop()

        if score > best_score:
            best_score = score
            best_move  = move
        if score > alpha:
            alpha = score

    return best_move, best_score
```

A few details worth noting:

- **Checkmate returns −10,000.** The side to move is in checkmate, so
  the score is a large negative number from their perspective. The
  parent negates this to +10,000 — a decisive win for the side that
  delivered mate.
- **`-beta, -alpha` in the recursive call.** The alpha and beta bounds
  are negated and swapped when passing to the child, because the child
  is evaluating from the opposite perspective.
- **Shared α across root moves.** In `get_best_move()`, `alpha` is
  updated as each root move is evaluated. This makes later root moves
  faster: any response worse than the current best is cut off
  immediately.

---

## Worked Example: One Position at Three Depths

We'll place the board after **1. e4 e5 2. Bc4 Nc6** — the setup for
Scholar's Mate. It's White's move. The engine will score every legal
move at depth 1 (look one half-move ahead) and depth 2 (look two
half-moves ahead, including Black's best response).

```{r load-weights}
con <- dbConnect(SQLite(), "../../data/chess_learning.db")

wt_rows <- dbGetQuery(con, "
    SELECT feature_name, weight_value
    FROM weights
    WHERE run_id = 12
      AND game_number = (SELECT MAX(game_number) FROM weights WHERE run_id = 12)
")
dbDisconnect(con)

weights_py <- as.list(setNames(wt_rows$weight_value, wt_rows$feature_name))
```

```{r setup-board}
board <- chess_m$Board()
for (uci in c("e2e4", "e7e5", "f1c4", "b8c6")) {
    board$push(chess_m$Move$from_uci(uci))
}

board  # position after 1. e4 e5 2. Bc4 Nc6
```

```{r score-moves}
# Score every legal move at a given depth.
# Each move is evaluated independently (no shared alpha across moves)
# so we see the true score for each option.
score_all_moves <- function(board, weights_py, depth) {
    moves <- reticulate::iterate(board$legal_moves)
    tibble(
        san   = purrr::map_chr(moves, \(m) board$san(m)),
        uci   = purrr::map_chr(moves, \(m) m$uci()),
        score = purrr::map_dbl(moves, \(m) {
            board$push(m)
            s <- -engine$negamax(board, as.integer(depth - 1), -1e6, 1e6, weights_py)
            board$pop()
            s
        })
    ) |> arrange(desc(score))
}

d1 <- score_all_moves(board, weights_py, depth = 1) |>
    rename(score_d1 = score)

d2 <- score_all_moves(board, weights_py, depth = 2) |>
    rename(score_d2 = score)

move_scores <- left_join(d1, d2, by = c("san", "uci")) |>
    mutate(
        rank_d1 = rank(-score_d1, ties.method = "first"),
        rank_d2 = rank(-score_d2, ties.method = "first"),
        rank_shift = rank_d1 - rank_d2   # positive = rose in ranking with lookahead
    ) |>
    arrange(rank_d2)
```

### Move Rankings: Depth 1 vs Depth 2

Each row is one legal White move. Scores are from White's perspective
(higher = better for White). **Rank shift** shows how a move's ranking
changes when Black's response is accounted for — a large positive value
means the move looks better with lookahead than without.

```{r display-table}
move_scores |>
    select(Move = san, `Score (d=1)` = score_d1, `Rank (d=1)` = rank_d1,
           `Score (d=2)` = score_d2, `Rank (d=2)` = rank_d2,
           `Rank shift` = rank_shift) |>
    flextable() |>
    colformat_double(j = c("Score (d=1)", "Score (d=2)"), digits = 3) |>
    color(
        j = "Rank shift",
        color = \(x) ifelse(x > 2, "darkgreen", ifelse(x < -2, "firebrick", "black"))
    ) |>
    set_table_properties(layout = "autofit")
```

### How Depth Changes the Best Move

```{r best-moves}
cat("Best move at depth 1:", move_scores |> slice_min(rank_d1, n=1) |> pull(san), "\n")
cat("Best move at depth 2:", move_scores |> slice_min(rank_d2, n=1) |> pull(san), "\n")

# Also run get_best_move at depth 3 — a single efficient call
best3 <- engine$get_best_move(board, weights_py, 3L, FALSE)
cat("Best move at depth 3:", best3[[1]]$uci(), "\n")
```

### Score Distribution

How spread out are the move scores? A wider spread means the engine has
a stronger preference between its options.

```{r score-plot}
#| fig-cap: "Each point is one legal move. Depth 1 (immediate evaluation only) vs Depth 2 (with Black's best response)."

move_scores |>
    select(san, score_d1, score_d2) |>
    pivot_longer(starts_with("score"), names_to = "depth", values_to = "score") |>
    mutate(depth = case_when(
        depth == "score_d1" ~ "Depth 1 — immediate only",
        depth == "score_d2" ~ "Depth 2 — with Black's response"
    )) |>
    ggplot(aes(x = score, y = depth)) +
    geom_jitter(height = 0.15, alpha = 0.6, size = 2) +
    labs(
        x = "Score (White's perspective)",
        y = NULL,
        title = "Move score distributions at depth 1 and depth 2"
    )
```

---

## What This Means for Learning

The search algorithm is fixed — it does not change across training runs.
What *does* change are the **weights** inside `evaluate()`. As training
progresses and the weights better reflect true chess value, the scores
produced by the evaluation function become more accurate, and
`get_best_move()` produces stronger play — not because the search
changed, but because the evaluations it backs up are better.

This separation of **search** (negamax) from **knowledge** (weights) is
the core architectural choice that makes the learning trajectories
interpretable: when a weight grows, we know exactly what concept the
engine has learned to value.
