---
title: "How the Chess Engine Learns"
subtitle: "A step-by-step walkthrough of one self-play game and its weight update"
author: "Posit AI, under guidance from Dan Killian"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    fig-height: 4
    fig-length: 6
execute:
  warning: false
  message: false
  include: true
  echo:    false
  error:   false
---

```{r}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
theme_set(base_quarto)

no_lead <- function(x, digits = 2) sub("^(-?)0\\.", "\\1.", as.character(round(x, digits)))
```

## Overview

This document walks through the engine's learning process in detail, using a single
short chess game as a concrete example. By the end you should understand:

- What a **weight** is and what it represents
- How the engine **evaluates** a board position using those weights
- How a game's result is used to **update** the weights via temporal difference learning
- Why some concepts (like queen value) are learned faster than others

---

## The evaluation function

The engine knows nothing about chess strategy on its own. Instead, it measures a
handful of features of any board position and combines them into a single score using
a weighted sum:

$$\text{score} = w_1 \times x_1 + w_2 \times x_2 + \cdots + w_n \times x_n$$

A positive score means the position looks good for the side to move; negative means
it looks bad.

In this tutorial we use three features to keep things readable. The real engine has
twelve.

```{r features}
feature_names <- c("material_queen", "king_safety", "center_control")

feature_descriptions <- tribble(
  ~Feature,           ~Description,
  "material_queen",   "Net queen advantage (my queens minus opponent's)",
  "king_safety",      "Composite king safety score (positive = safer)",
  "center_control",   "Control of the four central squares (positive = better)"
)

feature_descriptions |>
  flextable()
```

### Starting weights

The weights start as small random numbers. The engine has no prior knowledge —
it has to discover what matters entirely from the games it plays.

```{r initial-weights}
w <- tibble(feature=c("material_queen",
                     "king_safety",
                     "center_control"),
            label=c("Queen value", "King safety", "Center control"),
            weight=c(.02, -.03, .01)) # arbitrary
    
w |>
  flextable()
```

```{r plot-initial}

ggplot(w, aes(weight, label)) +
    geom_vline(xintercept=0, 
               color="grey60", 
               size=1,
               alpha=.5) +
    geom_label(aes(label=no_lead(weight, digits=4)),
               fill="indianred",
               color="white",
               fontface="bold",
               size=4) +
    scale_x_continuous(limits=c(-.05, .05),
                       breaks=seq(-.05,.05,.01),
                       labels = no_lead) +
  labs(title = "Initial weights",
       x = "Weight value",
       y = NULL) 
```

At this point, the engine evaluates almost every position as roughly equal — the
weights are too small to distinguish good positions from bad ones.

---

## The game: Scholar's Mate

We'll use Scholar's Mate — a four-move checkmate — as our example. It's short
enough to trace every step, and it has a decisive result (white wins), which
produces a clear learning signal.

The moves are: **1. e4 e5  2. Bc4 Nc6  3. Qh5 Nf6??  4. Qxf7#**

White wins by exploiting the weakness of black's f7 square. King safety is the
critical factor in this game — the engine should learn that from these four positions.

```{r import-game, include=FALSE}
embed_url <- "https://lichess.org/embed/game/9VP5REFK?theme=auto&bg=auto"
```

```{r show-game, include=TRUE}
htmltools::tags$iframe(
  src    = embed_url,
  width  = "100%",
  height = "450px",
  style  = "border: none;"
)
```

```{r positions, include=T}
# Each position is recorded just before a side makes their move.
# Feature values are from that side's perspective:
#   positive = good for the side to move, negative = bad.

positions <- tribble(
  ~ply, ~color,  ~material_queen, ~king_safety, ~center_control, ~comment,
  1,    "white",  0.00,            0.10,          0.15,  "After 1.e4 e5 — white to move",
  2,    "black",  0.00,           -0.25,          -0.10, "After 2.Bc4 — black to move\n(f7 under attack from black's view)",
  3,    "white",  0.00,            0.20,           0.10, "After 2...Nc6 — white to move",
  4,    "black",  0.00,           -0.75,          -0.10, "After 3...Nf6?? — black to move\n(king about to be mated)")

positions |>
#  select(ply, color, material_queen, king_safety, center_control) |>
  flextable() |>
  set_header_labels(
    ply            = "Ply",
    color          = "To move",
    material_queen = "material_queen",
    king_safety    = "king_safety",
    center_control = "center_control"
  ) |>
  add_footer_lines("Feature values at each position (from the side-to-move's perspective)")
```

::: {.callout-note}
## Why "from the side-to-move's perspective"?

The engine uses the **same** evaluation function for both colours. When it is
black's turn, a feature value of −0.75 for `king_safety` means *black's* king is
in danger. The sign of the feature value already encodes whose position is worse.
This lets one set of weights serve both sides symmetrically.
:::

---

## Evaluation after opening move

Let's have a board

```{r}
# Fresh board
b   <- chess_m$Board()
b
```
Let's start an object to collect the feature values at each half-move ('ply')

```{r}
feats <- tibble(feature=names(engine$default_weights("zero")),
                description = c(
                    "Pawn count", "Knight count", "Bishop count",
                    "Rook count", "Queen count",
                    "Pawns with no opposing pawns ahead on same/adjacent file",
                    "Extra pawns beyond one on any file (penalty)",
                    "Pawns with no friendly pawns on adjacent files (penalty)",
                    "Safe squares adjacent to king (not attacked by opponent)",
                    "Central squares (d4, d5, e4, e5) attacked",
                    "Rooks on files with no pawns of either color",
                    "Both rooks share a rank/file with no pieces between"),
                scoring = c(
                    rep("Our count minus theirs", 6),
                    rep("Negated — our minus theirs", 2),
                    rep("Our count minus theirs", 3),
                    "1/0 — our minus theirs"),
                start=0)

feats %>% 
    flextable()
```

Let's start an object to collect the weights and their trajectory each move

```{r}
set.seed(43)
wts <- tibble(criterion=names(engine$default_weights("zero")),
              start = round(rnorm(12, 0, .01), 4)) # random start to jumpstart learning
wts %>%
    flextable()
```

Starting position score is the weights times the feature values

```{r}
# State before ply 1 (starting position)
v_t <- sum(feats$start * wts$start)
v_t
```
### e4

```{r}
# Push 1.e4 and evaluate next state
b$push(chess_m$Move$from_uci("e2e4"))

feats$t1 <- as.numeric(reticulate::py_to_r(engine$extract_features(b)))

```

```{r}
feats_flx <- feats |>
    select(-1) |>
    flextable() |>
    height_all(height = 0.5, part = "body") |>
    hrule(rule = "exact", part = "body") |>
    set_table_properties(layout = "fixed") |>
    flextable::width(j = "description", width = 2) |>
    flextable::width(j = "scoring",     width = 1)

feats_flx
```

White attacks the center! Negative 1 for black. No value in occupying the center. 

New score

```{r}
v_t1 <- sum(feats$t1 * wts$start)
v_t1
```
New weights

```{r}
# TD(0) update: target is next-state value
lr    <- 0.001
error <- v_t1 - v_t

wts$w1    <- wts$start + lr * error * feats$start
```

```{r}
wts %>%
    flextable()
```

The weight for center control has not changed! This is due to the small learning weight. The engine needs to see that attack on d5 correlate with a win. 

## e5

```{r}
# Push 1.e4 and evaluate next state
b$push(chess_m$Move$from_uci("e7e5"))

feats$t2 <- as.numeric(reticulate::py_to_r(engine$extract_features(b)))
```

```{r}
v_t2 <- sum(w1 * feats_t2)
v_t2

# TD(0) update: target is next-state value
lr    <- 0.001
error <- v_t2 - v_t1

wts$w2    <- w1 + lr * error * feats_t2
 
```

```{r}
feats_flx <- feats |>
    select(-1) |>
    flextable() |>
    height_all(height = 0.5, part = "body") |>
    hrule(rule = "exact", part = "body") |>
    set_table_properties(layout = "fixed") |>
    flextable::width(j = "description", width = 2) |>
    flextable::width(j = "scoring",     width = 1)

feats_flx

wts %>%
    flextable()
```



## What the engine actually sees

The positions above use hand-crafted values to keep the arithmetic clear. Here we
run the actual feature extractor on each ply of Scholar's Mate and apply the same
temporal difference update — so you can see how the real numbers compare.

```{r engine-setup, include=FALSE}
engine  <- reticulate::import_from_path("engine", path = "../python")
chess_m <- reticulate::import("chess")
```

```{r real-game, include=FALSE}
scholar_moves <- c("e2e4", "e7e5", "f1c4", "b8c6", "d1h5", "g8f6", "h5f7")
move_san      <- c("1.e4", "1...e5", "2.Bc4", "2...Nc6",
                   "3.Qh5", "3...Nf6??", "4.Qxf7#")

lr_real  <- 0.001
wts      <- engine$default_weights("zero")
board    <- chess_m$Board()
rec_real <- list()

for (i in seq_along(scholar_moves)) {
  color  <- ifelse(reticulate::py_to_r(board$turn), "white", "black")
  target <- ifelse(color == "white", 1.0, -1.0)
  feats  <- as.list(reticulate::py_to_r(engine$extract_features(board)))
  pred   <- sum(unlist(wts) * unlist(feats))
  err    <- target - pred

  rec_real[[i]] <- tibble(
    ply            = i,
    move           = move_san[i],
    color          = color,
    king_safety    = feats[["king_safety"]],
    center_control = feats[["center_control"]],
    material_queen = feats[["material_queen"]],
    prediction     = round(pred, 4),
    target         = target,
    error          = round(err, 4),
    w_ks_before    = round(wts[["king_safety"]],    5),
    w_cc_before    = round(wts[["center_control"]], 5),
    w_mq_before    = round(wts[["material_queen"]], 5)
  )

  for (nm in names(wts)) {
    wts[[nm]] <- wts[[nm]] + lr_real * err * feats[[nm]]
  }

  board$push(chess_m$Move$from_uci(scholar_moves[i]))
}

game_real <- bind_rows(rec_real)
```

::: {layout="[55,45]"}
```{r show-real-board, include=TRUE}
htmltools::tags$iframe(
  src    = embed_url,
  width  = "100%",
  height = "450px",
  style  = "border: none;"
)
```

```{r real-weight-table, include=TRUE}
game_real |>
  flextable() |>
  set_header_labels(
    ply            = "Ply",
    move           = "Move",
    color          = "To move",
    king_safety    = "King safety",
    center_control = "Center control",
    material_queen = "Queen material",
    prediction     = "Prediction",
    target         = "Target",
    error          = "Error",
    w_ks_before    = "w king safety",
    w_cc_before    = "w center control",
    w_mq_before    = "w queen material"
  ) |>
  add_footer_lines("Weights shown are values before the update at that ply. All weights start at zero.")
```
:::

---

## The Temporal Difference Update, Position by Position

After the game ends — white wins — we work backwards through every position and
ask: *"given what I now know about how this game turned out, how wrong was my
evaluation at that point, and which weights were responsible?"*

The **target** for each position is the game result from that side's perspective:
- White to move and white wins → target = **+1**
- Black to move and white wins → target = **−1**

The **error** is `target − prediction`. A large error means the engine's evaluation
was badly wrong and the weights need a bigger nudge.

```{r td-update}
LEARNING_RATE <- 0.001
outcome       <- "white wins"

# Work through each position, updating weights in place
weights  <- setNames(w$weight, w$feature)    # start from initial weights
log_list <- list()

for (i in seq_len(nrow(positions))) {
  row    <- positions[i, ]
  feats  <- c(material_queen = row$material_queen,
              king_safety    = row$king_safety,
              center_control = row$center_control)
  target <- if (row$color == "white") 1.0 else -1.0

  prediction <- sum(weights[names(feats)] * feats)
  error      <- target - prediction
  nudges     <- LEARNING_RATE * error * feats

  log_list[[i]] <- tibble(
    ply        = row$ply,
    color      = row$color,
    target     = target,
    prediction = prediction,
    error      = error,
    nudge_mq   = nudges["material_queen"],
    nudge_ks   = nudges["king_safety"],
    nudge_cc   = nudges["center_control"],
    w_mq_after = weights["material_queen"] + nudges["material_queen"],
    w_ks_after = weights["king_safety"]    + nudges["king_safety"],
    w_cc_after = weights["center_control"] + nudges["center_control"]
  )

  weights[names(feats)] <- weights[names(feats)] + nudges
}

update_log <- bind_rows(log_list)
final_weights <- weights
```

```{r update-table, include=T}
update_log |>
  select(ply, color, prediction, target, error, nudge_mq, nudge_ks, nudge_cc) |>
  mutate(across(where(is.numeric), ~ round(., 5))) |>
  flextable() |>
  set_header_labels(
    ply        = "Ply",
    color      = "To move",
    prediction = "Prediction",
    target     = "Target",
    error      = "Error",
    nudge_mq   = "\u0394 material_queen",
    nudge_ks   = "\u0394 king_safety",
    nudge_cc   = "\u0394 center_control"
  ) |>
  add_footer_lines("Temporal difference update at each position — error drives the weight nudges")
```

A few things to notice:

- **Ply 1**: The engine predicted nearly 0 (weights are tiny). Target is +1 (white
  will win). Large error → noticeable nudges to `king_safety` and `center_control`.
- **Ply 2**: Black is to move, so target is −1. The engine again predicted nearly 0.
  Because `king_safety` is −0.25 (bad for black), the nudge pushes the king safety
  weight upward — a position where king safety is negative, combined with a loss,
  is consistent evidence that king safety matters.
- **Ply 4**: The error is largest here. `king_safety` is −0.75 (black king about
  to be mated) and the target is −1. The nudge to `king_safety` is the biggest in
  the game — this is the most informative position.
- **`material_queen` never changes**: it was 0.0 in all four positions (no queens
  were captured). A feature contributes to learning only when it varies.

---

## Weights before and after

```{r comparison-table}
out <- tibble(
  feature  = w$feature,
  label    = w$label,
  before   = w$weight,
  after    = final_weights,
  change   = final_weights - w$weight
) |>
  mutate(across(where(is.numeric), ~ round(., 6)))

out |>
  flextable() |>
  set_header_labels(
    feature = "Feature",
    before  = "Before",
    after   = "After",
    change  = "Change"
  ) |>
  add_footer_lines("Weight values before and after one game of Scholar's Mate")
```

```{r comparison-plot}

outL <- out |> 
    pivot_longer(c(before, after), names_to = "stage", values_to = "weight") 


ggplot(out, aes(x=change, y = label)) +
    geom_vline(xintercept=0, 
               color="grey60", 
               size=1,
               alpha=.5) +
    geom_label(aes(label=no_lead(change, digits=4)),
               fill="indianred",
               color="white",
               fontface="bold",
               size=4) +
    scale_x_continuous(limits=c(-.03, .03),
                       breaks=seq(-.03,.03,.01),
                       labels = no_lead) +
  labs(title = "Change in weights",
       x = "Weight value",
       y = NULL) 
```

The changes after a single game are tiny — that's intentional. The learning rate
(0.001) is deliberately small to prevent any one game from overwriting what was
learned from all previous games.

---

## What happens over many games

One game produces nudges too small to see in practice. The power of the approach
comes from repetition. Here is a simulated illustration: we run the same four
Scholar's Mate positions through 2,000 games, adding a small amount of noise to
the feature values each time to represent the natural variation between real games.

```{r simulation}
set.seed(3301)
N_GAMES    <- 2000
sim_w      <- setNames(w$weight, w$feature)    # restart from initial weights
trajectory <- tibble(game = 0, feature = names(sim_w), weight = sim_w)

for (g in seq_len(N_GAMES)) {
  for (i in seq_len(nrow(positions))) {
    row   <- positions[i, ]
    noise <- rnorm(3, sd = 0.15)
    feats <- c(material_queen = row$material_queen + noise[1],
               king_safety    = row$king_safety    + noise[2],
               center_control = row$center_control + noise[3])

    target     <- if (row$color == "white") 1.0 else -1.0
    prediction <- sum(sim_w[names(feats)] * feats)
    error      <- target - prediction
    sim_w[names(feats)] <- sim_w[names(feats)] + LEARNING_RATE * error * feats
    sim_w[] <- pmax(-50, pmin(50, sim_w))  # clip
  }
  if (g %% 10 == 0) {
    trajectory <- bind_rows(
      trajectory,
      tibble(game = g, feature = names(sim_w), weight = sim_w)
    )
  }
}
```

```{r}
#trajectory |>
#  mutate(feature = factor(feature, levels = feature_names)) 


ggplot(trajectory, aes(x = game, y = weight, colour = feature)) +
    geom_line(linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    labs(
        title    = "Simulated weight trajectories over 2,000 games",
        subtitle = "Noise added to feature values each game to simulate real variation",
        x        = "Game number",
        y        = "Weight value",
        colour   = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
```

`king_safety` climbs steadily toward a positive value because it consistently
distinguishes winning positions from losing ones. `center_control` also rises, but
more slowly — it is a weaker signal in Scholar's Mate. `material_queen` stays near
zero because it never varies in this particular game sequence; in a real training
run with many types of games, it would rise quickly as queen advantage is one of the
strongest predictors of winning.

::: {.callout-tip}
## The key intuition

Each weight is asking a question: *does this feature predict winning?*  
If yes, the weight grows. If the feature is irrelevant, it stays near zero.  
If the feature accidentally correlates with losing, it goes negative.  
After thousands of games, the weights reflect the true structure of chess.
:::
